Global code motion of congruent instructions

### 2017-11-06 cgo/paper.tex
corrections

We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
infrastructure available in \LLVM{}. The experimental results show an average
saving of 1\% on the total compilation time on \SPEC{}. \GCM{} enables more
inlining and exposes more loop invariant code motion opportunities in the
majority of benchmarks. We have also seen execution time improvements in a few
of \SPEC{} benchmarks viz. mcf and sjeng, moreover, register spills reduced by
2\% on the \SPEC{} benchmarks suite when compiled for \xlinux{}. \GCM{} is an
optimistic algorithm in the sense that it considers all congruent instructions
in a function as potential candidates. We make an extra effort to hoist
candidates by partitioning the potential candidates in a way to enable partial
hoisting in case a common hoisting point for all the candidates cannot be
found. We also formalize why register pressure reduces as a result of \gcm{},
and how \gcm{} increases instruction level parallelism thereby enabling more
\ooo{} execution on modern architectures.
\end{abstract}

@@

\section{A fast algorithm for code motion congruent computations}
The algorithm factors out the reachability of values such that multiple queries
to find reachability of values are fast. This is based on finding the dominance
frontiers and the post-dominance frontiers of the CFG. These points, being
structural properties of a graph, do not change during code-motion.

A post-dominance frontier is the basic block where anticipability (ANTIC) of an
instruction (I) is checked because it the point on which the basic block
containing I is control dependent.

So we introduce a data structure (CHI nodes) to keep track of values flowing out
of a basic block. We only do this for values with multiple occurrences in the
function as they are the potentially hoistable candidates. This approach allows
us to hoist instructions to a basic block with more than two successors, as well
as deal with infinite loops in a trivial way.

@@

A CHI keeps information about values flowing out of a basic block.  It is
similar to PHI but in the inverse graph, and used for outgoing values on each
edge. For conciseness, it is computed only for instructions with multiple
occurrences in the CFG because they are the only hoistable candidates.

\begin{figure}
  \includegraphics[width=0.4\textwidth]{chi-example.pdf}
  \vspace*{-1.0cm}
\caption{I1, I2 are instructions in B, and C respectively, CHIArg1 = \{V, B,
  I1\} and CHIArg2 = \{V, C, I2\}, V is the value number for both I1, and I2}
\label{fig:chi-intro}
\end{figure}

The global value number for both I1 and I2 is V, the CHI node will save the
instruction as well as the edge where the value is flowing to.  Where CHIArg1,
CHIArg2 are of type CHIArg.

Essentially, CHIArg1 = \{V, B, I1\} and CHIArg2 = \{V, C, I2\}. For sinking we
use the similar approach as \cite{ssapre} to track values flowing into a basic
block.

@@

\begin{itemize}
\item Sort VNs according to their rank, and start with lowest ranked VN
\item Take a VN and for each instruction with same VN
\item  - Find the dominance frontier in the inverse graph (PDF)
\item  - Insert the chi-node at PDF
\item Remove the chi-nodes with missing entries
\item Remove values from CHI-nodes which do not truly flow out, e.g.,
\item  modified along the path.
\item Collect the remaining values that are still anticipable
\end{itemize}

@@

\begin{verbatim}
B0: b = m
    c = n
    if c is true then goto B1 else goto B2
B1: a0 = b<kill> + c<kill>
    goto B3
B2: a1 = b<kill> + c<kill>
    goto B3
\end{verbatim}

@@

There is a special case for hoisting load instructions when the hoist-point is
the predecessor basic-block for all the congruent loads. Even if the register
pressure would increase, we prefer to hoist loads in this case. The reason
being, that would make the loaded value available by the time it is used.

@@

\subsection{Finding candidates to move}
\label{subsec:finding-candidates}
The first step is to find a set of congruent instructions
\cite{briggs1997}. This is performed by a linear scan of all instructions of the
program and classifying them by their value numbers. We could compute available
and anticipable sets as computed by \GCC{'}s code-hoisting but that would be a lot
of data structures to maintain at each basic-block level. For \GCC{} it makes sense
because their code-hoisting is integrated with \GVN{}-\PRE{} which already has those
data structures available.

The current implementation of \GVN{} in \LLVM{} has some limitations when it
comes to loads and stores so we compute the \GVN{} of loads and stores
separately.  Our solution is to value number the address from where the value of
a load is to be read from memory. For stores, we value number the address as
well as the value to be stored at that address. Another limitation of the
current \GVN{} implementation in \LLVM{} is that the instructions dependent on
the loads will not get numbered correctly, and so after hoisting all candidates
we need to rerun the \GVN{} analysis in order to discover new candidates now
available after having hoisted load instructions.  This limitation should be
addressed in a new implementation of the \GVN{} based on \MemorySSA{}, that
would better account for equivalent loads and their dependent instructions.

@@

Once the instructions have been classified into congruence classes, we compute
for each group of congruent instructions, a point in the program that is both
legal and profitable for the instructions to be moved to.

@@

\subsubsection{Partitioning the list of hoist candidates to maximize hoisting}
\label{subsec:partition}
%The partition algorithm is described as follows:
In the approach described in Section~\ref{subsec:optimistic}, it is possible
that a common hoisting point for all the instructions does not satisfy legality
or profitability checks. In these cases, it is still possible to `partially'
hoist a subset of instructions by splitting the set of candidates and finding a
closer hoisting point for each subset.

In order to hoist a subset of identical instructions, we partition the list of
all candidates in a way to maximize the total number of hoist instructions.  By
sorting the list of all the candidates in the increasing order of their depth
first search discovery time stamp \cite{clrs} (DFSIn numbers) we make sure that
candidates closer in the list have their common dominator nearby. In
Figure-\ref{fig:dfsin}, if B3, B4, B5, and B6 have identical computations and if
for some reason, they cannot be hoisted at B0, then we partition the set of
hoistable candidates in their DFSIn order. In this case the DFSIn ordering would
be B3, B4, B5, B6 which will allow the instructions in (B3, B4) to be hoisted at
B1 and those in (B5, B6) to be hoisted at B2. Even if instructions in (B3, B4)
could not be hoisted, for some reason, to B1, the instructions in (B5, B6) can
still be hoisted - if legal and profitable - to B2.

\begin{figure}
\centering
%rankdir=LR;
\digraph[scale=0.35]{abc}{
  B0 -> B1
  B0 -> B2
  B1 -> B3
  B1 -> B4
  B2 -> B5
  B2 -> B6
}
\caption{\CFG{} to illustrate partitioning}
\label{fig:dfsin}
\end{figure}

% TODO: Rewrite the following.
In our current implementation we keep as many candidates in one set as possible
(greedy approach.) To start with, the first two candidates are tried for
hoistability (see Section~\ref{subsec:optimistic}) if that passes then the next
instruction is tried (if available) and so on until a non-hoistable instruction
is found. The partition starts from that point.  We split the list at a point
where the legality checks fail to hoist subset of candidates which are legal to
hoist and then start finding new hoisting point for the remaining ones.

\subsection{Legality check}
\label{subsec:legality}
Since the equality of candidates is purely based on the value numbers, we also
need to establish if hoisting them to a common dominator or sinking them to a
common post-dominator would be legal. Once a common dominator (post-dominator)
is found, we check whether all the use-operands of the set of instructions are
available at that position. It is possible to reinstantiate or remateralize the
use-operands in some cases when the operands are not available and make it legal
to move the instruction.

Subsequently, it is checked that the side-effects of the computations (if any)
do not intersect with any side-effects between the instructions to be
moved and their hoisting/sinking point. It is also necessary to check if
there are indirect branch targets, e.g., landing pad, case statements, goto
labels, etc., along the path because it becomes difficult to prove safety checks
in those cases. In our current implementation we discard candidates on those
paths.

@@

\subsubsection{Legality of hoisting loads}
The availability of operand to the load (an address) is checked at the hoisting
point. If that is not available we try to rematerialize the address if
possible.  Along the path, from current position of the load instruction
backwards on the control flow to the hoisting point, we check whether there are
writes to memory that may alias with the load, in which case the candidate is
discarded. To iterate on the use-def chains for memory references the MemorySSA
of \LLVM{} is used.

\subsubsection{Legality of hoisting stores}
For stores, we check the dependency requirements similar to the hoisting of
loads using the \MemorySSA{} of \LLVM{}. We check that the operands of the store
instruction are available at the hoisting point, and there are no aliasing
memory references along the hoisting path.

\subsubsection{Legality of hoisting calls}
Call instructions can be divided into three categories: those calls equivalent
to purely scalar computations, calls reading from memory, and most of the time,
without further information, calls have to be classified as writing to memory,
that is the most restrictive form.  Each category of call instructions is
handled as described for scalar, load, and store instructions.

Hoisting loads/stores across calls also requires precise analysis of all the
memory addresses accessed by the call. The current implementation being an
intraprocedural pass, cannot hoist aggressively across calls. In the presence of
pure calls, loads can be hoisted but stores can't. Also, if a call throws
exceptions, or if it may not return, nothing can be hoisted across that call.

@@

In order to be able to move an instruction that has side effects, we must
guarantee that the side-effects of that instruction before and after its
code-motion will produce the same side-effects.  For hoisting in particular, we
need to prove that on all execution paths from the hoisting point to the end of
the program the side effects appear in the same order. In general, we cannot
introduce a new computation along any path of execution without checking for
undefined behaviors. The algorithm \emph{computeDownwardSafety} computes if
hoisting a set of instruction to a common hoisting point is downward-safe
\cite{steven1997advanced}. Since \GCM{} is only dealing with VBEs, computing
downward-safety ensures that no undefined behaviors are introduced as a result
of code-motion.
\newline
\begin{verbatim}
// Return true if a hoistable instruction is downward
// safe at hoist-point
bool computeDownwardSafety(Node HP)
  Worklist = basic-blocks of hoistable instructions
  DT = The dominator tree
  HP = the hoist-point
  for each basic-block B in the DT starting at HP:
    if Worklist is empty
       return false // Path exists! not downward safe
    if B is in Worklist:
      remove B from worklist
      // Available at B => downward safe from B
      remove subtree with root at B
    if B is a leaf basic-block
      // Path exists! not downward safe
      return false
    if B dominates hoist-point
      return false  // Back edge
  return true
\end{verbatim}

@@

\begin{verbatim}
void doGCM()
  Analyses available:
    Dominator Tree, DFS Numbering, Memory SSA
  Compute DJ-graph of function
  constructMergeSet(CGF)
  computeBarriers()
  do:
    Compute GVN of each expression in the function
   // Repeat hoisting if any candidates were hoisted.
  while (doCodeHoisting() > 0)
  doCodeSinking()
\end{verbatim}

@@

For hoisting, we find a common dominator dominating all a set of congruent
instructions and perform legality checks, as described in
Section~\ref{subsec:legality}. Often times it is not possible to hoist all the
instructions to one common dominator, due to legality constraints, e.g.,
intersecting side-effects, or profitability constraints, e.g., increasing
register pressure; in those cases, this algorithm would partition
(Section~\ref{subsec:partition}) the list of identical instructions into subsets
which can be partially hoisted to their respective common dominators.

\begin{verbatim}
// Find hoistable candidates and do hoisting
int doCodeHoisting()
  For each value number VN with 2 or more instructions
    sort the instructions according to DFSIn numbers
    if first two I1, I2 are hoistable
        && downward-safe at hoist-point
        && profitable to be hoisted at hoist-point:
      continue the check for subsequent
        instructions with same VN
    else:
      proceed to check if I2 and I3 (if available)
      can be hoisted

  For each set of hoistable instructions:
    move the first one (I1) to the hoist-point
    update their users others to refer to I1
    remove all other instructions (I2, ..., Ik)
    if I1 has memory references:
      update MemorySSA of I1 and others
    update statistics
  return number of hoisted instructions
\end{verbatim}

>>>>

We have implemented the pass in \LLVM{}. It is based on Global Value Numbering
infrastructure available in \LLVM{}. The experimental results show an average
saving of 1\% on the total compilation time on \SPEC{}. \GCM{} enables more
inlining and exposes more loop invariant code motion opportunities in the
majority of benchmarks. We have also seen execution time improvements in a few
of \SPEC{} benchmarks viz. mcf and sjeng, moreover, register spills reduced by
2\% on the \SPEC{} benchmarks suite when compiled for \xlinux{}. \GCM{} is an
optimistic algorithm in the sense that it considers all congruent instructions
in a function as potential candidates. We also formalize why register pressure
reduces as a result of \gcm{}, and how \gcm{} increases instruction level
parallelism thereby enabling more \ooo{} execution on modern architectures.
\end{abstract}

@@

\section{A fast algorithm for code motion congruent computations}
The algorithm factors out the reachability of values such that multiple queries
to find reachability of values are fast. This is based on finding the dominance
frontiers and the post-dominance frontiers (PDF) of the CFG. These points, being
structural properties of a graph, do not change during code-motion. A
post-dominance frontier is the basic block where anticipability (ANTIC) of an
instruction (I) is checked because it the point on which the basic block
containing I is control dependent.

So we introduce a data structure (CHI nodes) to keep track of values flowing out
of a basic block with more than once successors. We only do this for values with
multiple occurrences in the function as they are the potentially hoistable
candidates. This approach allows us to hoist instructions to a basic block with
more than two successors. The CHI node in a basic block $B$ has CHIArgs with
three entries: a) the instruction (I) CHIArg refers to, b) the basic block $Dest$
which is post-dominated by the basic block containing I, and c) the GVN of I.

@@

A CHI keeps information about values flowing out of a basic block.  It is
similar to PHI but in the inverse graph, and used for outgoing values on each
edge. Although it may be suitable for other analyses/optimizations, for \GCM{},
it is computed only for instructions with multiple occurrences in the CFG. An
illustration of CHI is shown in Figure~\ref{fig:chi-intro}. The \GVN{} for both
I1 and I2 is V, the CHI node will save the instruction as well as the edge where
the value is flowing to.  Where CHIArg1, CHIArg2 are of type CHIArg. The CHI for
both I1 and I2 is inserted in their PDF. For sinking we use the similar approach
as \cite{ssapre} to track values flowing into a basic block.

\begin{figure}
  \includegraphics[width=0.4\textwidth]{chi-example.pdf}
  \vspace*{-1.0cm}
\caption{I1, I2 are instructions in B, and C respectively, CHIArg1 = \{B, I1,
  V\} and CHIArg2 = \{C, I2, V\}, V is the value number for both I1, and I2}
\label{fig:chi-intro}
\end{figure}

@@

@@

\begin{verbatim}
B0: b = m
    c = n
    if p is true then goto B1 else goto B2
B1: a0 = b<kill> + c<kill>
    goto B3
B2: a1 = b<kill> + c<kill>
    goto B3
B3: ...
\end{verbatim}

@@

There is a special case for hoisting load instructions when the hoist-point is
the predecessor basic-block for all the congruent loads. Even if the register
pressure would increase, we prefer to hoist loads in this case. The reason
being, that would make the loaded value available by the time it is used, and
livess would not increase by much because instruction is hoisted to immediate
predecessor.

@@

\subsection{Finding candidates to move}
\label{subsec:finding-candidates}
The first step is to find a set of congruent instructions
\cite{briggs1997}. This is performed by a linear scan of all instructions of the
program and classifying them by their value numbers.

The current implementation of \GVN{} in \LLVM{} has some limitations when it
comes to loads and stores so we compute the \GVN{} of loads and stores
separately.  Our solution is to value number the address from where the value of
a load is to be read from memory. For stores, we value number the address as
well as the value to be stored at that address. Another limitation of the
current \GVN{} implementation in \LLVM{} is that the instructions dependent on
the loads will not get numbered correctly, and so after hoisting all candidates
we need to rerun the \GVN{} analysis in order to discover new candidates now
available after having hoisted load instructions.

@@

@@

\subsection{Legality check}
\label{subsec:legality}
Since the equality of candidates is purely based on the value numbers, we also
need to establish if hoisting them or sinking them would be legal. Once a common
dominator (post-dominator) is found, we check whether all the use-operands of
the set of instructions are available at that position. It is possible to
reinstantiate or remateralize the use-operands in some cases when the operands
are not available and make it legal to move the instruction.

Subsequently, it is checked that the side-effects of the computations (if any)
do not intersect (alias) with any side-effects between the instructions to be
moved and their hoisting/sinking point. It is also necessary to check if there
are indirect branch targets, e.g., landing pad, case statements, goto labels,
etc., along the path because it becomes difficult to prove safety checks in
those cases. In our current implementation we discard candidates on those paths.

@@

\subsubsection{Legality of hoisting memory references}
The availability of operand to the load (an address) is checked at the hoisting
point. If that is not available we try to rematerialize the address if possible.
Along the path, from current position of the load instruction backwards on the
control flow to the hoisting point, we check whether there are writes to memory
that may alias with the load, in which case the candidate is discarded. Memory
SSA allows efficient way of iterating over the use-def chains of memory
references on a factored graph.

For stores, we check the dependency requirements similar to the hoisting of
loads. We check that the operands of the store instruction are available at the
hoisting point, and there are no aliasing memory references along the hoisting
path.

\subsubsection{Legality of hoisting calls}
Call instructions can be divided into three categories: those calls equivalent
to purely scalar computations, calls reading from memory, and calls writing to
memory which is the most restrictive form.  Each category of call instructions
is handled as described for scalar, load, and store instructions.

Hoisting loads/stores across calls also requires precise analysis of all the
memory addresses accessed by the call. The current implementation being an
intraprocedural pass, cannot schedule aggressively across calls. In the presence
of pure calls, loads can be hoisted but stores can't. Also, if a call throws
exceptions, or if it may not return, nothing can be scheduled across that call.

@@

In order to be able to move an instruction that has side effects, we must
guarantee that the side-effects of that instruction before and after its
code-motion will produce the same side-effects.  For hoisting in particular, we
need to prove that on all execution paths from the hoisting point to the end of
the program the side effects appear in the same order. In general, we cannot
introduce a new computation along any path of execution without checking for
undefined behaviors \cite{undef}. Since \GCM{} is only dealing with VBEs, as
hoisting happens for CHIs with complete entries, the downward safety is ensured
by construction of CHIs.

@@

\begin{verbatim}
void doGCM()
  Analyses available:
    Dominator Tree, DFS Numbering, Memory SSA
  Compute DJ-graph of function
  constructMergeSet(CFG)
  computeBarriers()
  do:
    Compute GVN of each expression in the function
   // Repeat hoisting if any candidates were hoisted.
  while (doCodeHoisting() > 0)
  doCodeSinking()
\end{verbatim}

@@

\begin{verbatim}
int doCodeHoisting()
  Sort VNs in ascending order

  for each instruction with the same VN
      Find its post-dominance-frontier (PDF)
      Insert the CHI-node at PDF

  for each CHI-nodes in the CFG:
    Remove the ones with missing entries
    Remove CHI-arg if it fails legality checks

  for each CHI-nodes in the CFG:
    if CHI  does not have missing entries:
      Perform code hoisting
